---
title: "damidBind manuscript figures and tables"
author: "Owen Marshall"
output: html_notebook
---

This file is an R Notebook of all of the analysis figures and tables from Marshall (2026) "damidBind: an R/Bioconductor package for differential DamID analysis and data exploration", with the exception of two larger multi-panel supplementary figures.

The code for those figures, Fig. S5 and Fig. S7, is provided as R scripts separately within this archive.

```{r}
pacman::p_load(damidBind,dplyr,ggplot2)
```
 
```{r}
#' Benchmark damidBind results against Xu et al 2024 bespoke analysis and 
#' scRNA-seq markers as a form of ground truth
#'
#' @param results_obj A \code{DamIDResults} object.
#' @param bespoke_obj Object provided by previous manual/bespoke analysis 
#'   (expected to have \code{$upreg} and \code{$downreg} vectors).
#' @param ground_truth_genes Character vector of gene symbols (e.g. scRNA-seq markers).
#' @param target_cond Integer (1 or 2) indicating which condition to benchmark.
#' @param label String. A label for the results (e.g., "L4 Neurons").
#'
#' @return A list containing:
#'   \itemize{
#'     \item \code{locus_stats}: Dataframe of Fisher's tests at the locus level.
#'     \item \code{gene_stats}: Dataframe of Fisher's tests at the gene level.
#'     \item \code{recovery_stats}: Dataframe showing percent recovery of bespoke 
#'       results by the package.
#'     \item \code{volcano_plot_all}: The volcano plot of the \code{damidBind} object, 
#'       with significant bespoke hits from both conditions overlaid as highlights.
#'     \item \code{volcano_plot_cond}: The volcano plot showing package hits, 
#'       bespoke hits for the target condition, and ground truth markers.
#'   }
#' @export
benchmark_results <- function(results_obj, 
                              bespoke_obj, 
                              ground_truth_genes, 
                              target_cond = 1, 
                              label = "Cell Type") {
    
    # Validation and data extraction
    if (!inherits(results_obj, "DamIDResults")) {
        stop("results_obj must be a DamIDResults object.")
    }
    
    analysis_tab <- analysisTable(results_obj) %>%
        tibble::rownames_to_column("locus")
    
    # Define significant loci for the package for the target condition
    if (target_cond == 1) {
        package_loci <- rownames(enrichedCond1(results_obj))
        bespoke_loci <- bespoke_obj$upreg
    } else {
        package_loci <- rownames(enrichedCond2(results_obj))
        bespoke_loci <- bespoke_obj$downreg
    }
    
    # Map Loci to Genes
    gene_locus_map <- analysis_tab %>%
        dplyr::select(locus, gene_name) %>%
        dplyr::filter(gene_name != "" & !is.na(gene_name)) %>%
        tidyr::separate_rows(gene_name, sep = ",") %>%
        dplyr::mutate(gene_name = trimws(gene_name))
    
    gene_universe <- unique(gene_locus_map$gene_name)
    ground_truth_universe <- intersect(ground_truth_genes, gene_universe)
    
    # Locus-level Enrichment Test
    gt_loci <- gene_locus_map %>%
        dplyr::filter(gene_name %in% ground_truth_universe) %>%
        dplyr::pull(locus) %>%
        unique()
    
    analysis_tab <- analysis_tab %>%
        dplyr::mutate(
            is_gt = locus %in% gt_loci,
            is_package = locus %in% package_loci,
            is_bespoke = locus %in% bespoke_loci
        )
    
    calc_locus_fet <- function(df, target_col, sub_label) {
        # a: Sig + GT; b: Sig + Not GT; c: Not Sig + GT; d: Not Sig + Not GT
        a <- sum(df[[target_col]] & df$is_gt)
        b <- sum(df[[target_col]] & !df$is_gt)
        c <- sum(!df[[target_col]] & df$is_gt)
        d <- sum(!df[[target_col]] & !df$is_gt)
        
        ft <- fisher.test(matrix(c(a, c, b, d), nrow = 2))
        
        data.frame(
            comparison = sub_label,
            total_detected = sum(df[[target_col]]),
            gt_overlap = a,
            percent_overlap = (a / sum(df[[target_col]])) * 100,
            odds_ratio = unname(ft$estimate),
            p_value = ft$p.value
        )
    }
    
    locus_res <- dplyr::bind_rows(
        calc_locus_fet(analysis_tab, "is_package", paste(label, "(Package Total)")),
        calc_locus_fet(analysis_tab, "is_bespoke", paste(label, "(Xu et al Total)")),
        calc_locus_fet(dplyr::filter(analysis_tab, !is_bespoke), "is_package", paste(label, "(Novel Only)"))
    )
    
    # Gene-level Enrichment Test
    package_genes <- gene_locus_map %>%
        dplyr::filter(locus %in% package_loci) %>%
        dplyr::pull(gene_name) %>%
        unique()
    
    bespoke_genes <- gene_locus_map %>%
        dplyr::filter(locus %in% bespoke_loci) %>%
        dplyr::pull(gene_name) %>%
        unique()
    
    calc_gene_fet <- function(test_genes, universe, ground_truth, sub_label) {
        a <- length(intersect(test_genes, ground_truth))
        b <- length(setdiff(test_genes, ground_truth))
        c <- length(setdiff(ground_truth, test_genes))
        d <- length(setdiff(universe, dplyr::union(test_genes, ground_truth)))
        
        ft <- fisher.test(matrix(c(a, c, b, d), nrow = 2))
        
        data.frame(
            comparison = sub_label,
            total_genes = length(test_genes),
            gt_overlap = a,
            percent_overlap = (a / length(test_genes)) * 100,
            odds_ratio = unname(ft$estimate),
            p_value = ft$p.value
        )
    }
    
    gene_res <- dplyr::bind_rows(
        calc_gene_fet(package_genes, gene_universe, ground_truth_universe, paste(label, "(Package Total)")),
        calc_gene_fet(bespoke_genes, gene_universe, ground_truth_universe, paste(label, "(Xu et al Total)")),
        calc_gene_fet(
            setdiff(package_genes, bespoke_genes), 
            setdiff(gene_universe, bespoke_genes), 
            setdiff(ground_truth_universe, bespoke_genes), 
            paste(label, "(Novel Only)")
        )
    )
    
    locus_recovery <- length(intersect(package_loci, bespoke_loci))
    gene_recovery <- length(intersect(package_genes, bespoke_genes))
    
    rec_stats <- data.frame(
        level = c(sprintf("%s (locus)",label), sprintf("%s (gene)",label)),
        bespoke_total = c(length(bespoke_loci), length(bespoke_genes)),
        recovered = c(locus_recovery, gene_recovery),
        percent_recovery = c(
            (locus_recovery / length(bespoke_loci)) * 100,
            (gene_recovery / length(bespoke_genes)) * 100
        )
    )
    
    # Visualisations 
    # Global comparison
    comp_plot <- plot_volcano(
        results_obj,
        highlight = list(
            "L4 (Xu et al)" = bespoke_obj$upreg,
            "L5 (Xu et al)" = bespoke_obj$downreg
        ),
        labels = "all",
        highlight_config = list(
            alpha = 0.5,
            size = 0.8,
            legend_inside_pos = "l"
        ),
    ) + ggplot2::labs(title = label)
    
    # Condition-specific comparison with scRNA-seq
    v_plot <- plot_volcano(
        results_obj,
        highlight = list(
            "Xu et al 2024" = bespoke_loci,
            "scRNA-seq" = ground_truth_universe
        ),
        labels = "scRNA-seq",
        highlight_config = list(
            alpha = 0.5,
            size = 0.8,
            sig_labels_only = TRUE,
            legend_inside_pos = "l"
        )
    ) + ggplot2::labs(title = label)
    
    return(list(
        locus_stats = locus_res,
        gene_stats = gene_res,
        recovery_stats = rec_stats,
        volcano_plot_all = comp_plot,
        volcano_plot_cond = v_plot
    ))
}

```

## Analysis

### Load up Xu et al 2024 analysis
```{r}
# load the pre-saved bespoke Bsh analysis:
load("data/Xu_et_al_2024_data.RData")

# scRNA-seq analysed subtype markers from the paper
l4_spec_mar = read.csv("data/scRNA-seq_markers/l4_specific_March.csv",header=T,row.names = 1)
names(l4_spec_mar)[1]="name"
l5_spec_mar = read.csv("data/scRNA-seq_markers/l5_specific_March.csv",header=T,row.names = 1)
names(l5_spec_mar)[1]="name"

l4_rnaseq_spec_mar = capture.output(cat(l4_spec_mar$name,sep = "|"))
l5_rnaseq_spec_mar = capture.output(cat(l5_spec_mar$name,sep = "|"))
```


### Fig S3 is a composite of the Limma diagnostic plots from the following:
```{r load_data}
# Perform the damidBind Bsh TaDa analysis
# We use the same paths and settings as the original manuscript
xubsh_loess <- load_data_peaks(
    binding_profiles_path = "data/Bsh/*ph.gz",
    peaks_path = "data/Bsh/*gff.gz",
    norm_method = "loess",
    plot_diagnostics = TRUE
)

xubsh_qnorm <- load_data_peaks(
    binding_profiles_path = "data/Bsh/*ph.gz",
    peaks_path = "data/Bsh/*gff.gz",
    norm_method = "quantile",
    plot_diagnostics = TRUE
)

xubsh_none <- load_data_peaks(
    binding_profiles_path = "data/Bsh/*ph.gz",
    peaks_path = "data/Bsh/*gff.gz",
    norm_method = "none",
    plot_diagnostics = TRUE
)


png("Fig_S3_%03d.png",width=(6.5*200), height=(5.5*200), res=200)

xubsh_diff_loess <- differential_binding(
    xubsh_loess,
    cond = c("L4 neurons" = "L4", "L5 neurons" = "L5"),
    plot_diagnostics = TRUE
)

xubsh_diff_qnorm <- differential_binding(
    xubsh_qnorm,
    cond = c("L4 neurons" = "L4", "L5 neurons" = "L5"),
    plot_diagnostics = TRUE
)

xubsh_diff_none <- differential_binding(
    xubsh_none,
    cond = c("L4 neurons" = "L4", "L5 neurons" = "L5"),
    plot_diagnostics = TRUE
)

dev.off()
```

### Run through the benchmark comparisons ...
```{r}
# scaled only
none <- benchmark_results(
    results_obj = xubsh_diff_none,
    bespoke_obj = bespoke_bsh_results,
    ground_truth_genes = l4_spec_mar$name,
    target_cond = 1,
    label = "L4 scaled-only"
)

none
```

### Fig. S6(A)
```{r}
# Quantile
qnorm <- benchmark_results(
    results_obj = xubsh_diff_qnorm,
    bespoke_obj = bespoke_bsh_results,
    ground_truth_genes = l4_spec_mar$name,
    target_cond = 1,
    label = "L4 neurons"
)

qnorm

png("Fig_S6A_L4.png",width=6*300, height=4*300,res=300)
qnorm$volcano_plot_cond
dev.off()
```

```{r}
# loess
loess <- benchmark_results(
    results_obj = xubsh_diff_loess,
    bespoke_obj = bespoke_bsh_results,
    ground_truth_genes = l4_spec_mar$name,
    target_cond = 1,
    label = "L4 loess norm"
)

loess
```

```{r}
none_l5 <- benchmark_results(
    results_obj = xubsh_diff_none,
    bespoke_obj = bespoke_bsh_results,
    ground_truth_genes = l5_spec_mar$name,
    target_cond = 2,
    label = "L5 scaled only"
)

none_l5
```

### Fig. S6(A)
```{r}
qnorm_l5 <- benchmark_results(
    results_obj = xubsh_diff_qnorm,
    bespoke_obj = bespoke_bsh_results,
    ground_truth_genes = l5_spec_mar$name,
    target_cond = 2,
    label = "L5 neurons"
)

qnorm_l5

png("Fig_S6A_L5.png",width=6*300, height=4*300,res=300)
qnorm_l5$volcano_plot_cond
dev.off()
```

```{r}
# loess
loess_l5 <- benchmark_results(
    results_obj = xubsh_diff_loess,
    bespoke_obj = bespoke_bsh_results,
    ground_truth_genes = l5_spec_mar$name,
    target_cond = 2,
    label = "L5 loess norm"
)

loess_l5
```

### Table S4, Table S5
```{r}
table_out = do.call(rbind,list(qnorm$locus_stats,qnorm_l5$locus_stats,
                               loess$locus_stats,loess_l5$locus_stats)
                    )
table_out
table_out <- table_out %>%
  mutate(p_value = formatC(p_value, format = "e", digits = 2))
knitr::kable(table_out,digits=2) %>% kableExtra::kable_styling()
```

### Table S6 
```{r}
table_out_recovery = do.call(rbind,list(qnorm$recovery_stats,qnorm_l5$recovery_stats,loess$recovery_stats,loess_l5$recovery_stats))
table_out_recovery
knitr::kable(table_out_recovery,digits=2) %>% kableExtra::kable_styling()
```



### Figure S4
```{r}
xucat_none = load_data_peaks(
    binding_profiles_path = "data/catada/*ph.gz", 
    peaks_path = "data/catada/*gff.gz",
    pre_scale=FALSE, 
    norm_method = "none")

xucat_diff_none = differential_accessibility(xucat_none,cond=c("L4 neurons"="L4","L5 neurons"="L5"))

mean_variance = plot_catada_mean_variance(xucat_diff_none)

ggsave("Fig_S4.pdf",mean_variance,width=6,height=3.5)
```


```{r}

xucat_qnorm =  load_data_peaks(
    binding_profiles_path = "data/catada/*ph.gz", 
    peaks_path = "data/catada/*gff.gz",
    pre_scale=FALSE, 
    norm_method = "quantile")

xucat_diff = differential_accessibility(xucat_qnorm,cond=c("L4 neurons"="L4","L5 neurons"="L5"))
```


### Fig S6(B)
```{r}
catada_l4 <- benchmark_results(
    results_obj = xucat_diff,
    bespoke_obj = bespoke_catada_results,
    ground_truth_genes = l4_spec_mar$name,
    target_cond = 1,
    label = "L4 neurons"
)

catada_l4

png("Fig_S6B_L4.png",width=6*300, height=4*300,res=300)
catada_l4$volcano_plot_cond
dev.off()
```

```{r}
catada_l5 <- benchmark_results(
    results_obj = xucat_diff,
    bespoke_obj = bespoke_catada_results,
    ground_truth_genes = l5_spec_mar$name,
    target_cond = 2,
    label = "L5 neurons"
)

catada_l5

png("Fig_S6B_L5.png",width=6*300, height=4*300,res=300)
catada_l5$volcano_plot_cond
dev.off()
```

### Table S8 and Table S9
```{r}
table_out.catada = do.call(rbind,list(catada_l4$locus_stats, catada_l5$locus_stats))
table_out.catada
table_out.catada <- table_out.catada %>%
  mutate(p_value = formatC(p_value, format = "e", digits = 2))
knitr::kable(table_out.catada,digits=2) %>% kableExtra::kable_styling()

table_out_recovery.catada = do.call(rbind,list(catada_l4$recovery_stats,catada_l5$recovery_stats))
table_out_recovery.catada
knitr::kable(table_out_recovery.catada,digits=2) %>% kableExtra::kable_styling()
```


### Figure 1B panel
```{r}
plot_volcano(xucat_diff,
             label_config=NULL,
             highlight=list(
                 "L4 specific"=l4_spec_mar$name,
                 "L5 specific"=l5_spec_mar$name
                 ),
             label_display = list(r=0.25),
             highlight_config = list(
                 legend_inside_pos='l',
                 text_col=T,
                 size=1,
                 text_luminosity=40, 
                 sig_labels_only=T,
                 max_overlaps=8
                 ),
             save = list(
                 filename="Fig_1B_CATaDa",
                 format="svg",
                 width=6,
                 height=4
                )
)
```

### Figure 1B panel
```{r}
plot_volcano(xubsh_diff_qnorm,
             label_config=NULL,
             highlight=list(
                 "L4 specific"=l4_spec_mar$name,
                 "L5 specific"=l5_spec_mar$name
                 ),
             label_display = list(r=0.25),
             highlight_config = list(
                 legend_inside_pos='l',
                 text_col=T,
                 text_luminosity=40, 
                 sig_labels_only=T,
                 size=1,
                 max_overlaps=8
                 ),
             save = list(
                 filename="Fig_1B_Bsh",
                 format="svg",
                 width=6,
                 height=4
                )
             )
```


### Figure S2
```{r}
bsh_l5_bias_q = test_weighting_vs_bias_artifact(xubsh_diff_qnorm,"Bsh_Dam_L5_r1-ext300-vs-Dam.kde-norm_qnorm")

png("Fig_S2.png", width = 3000, height = 1500, res = 300)
bsh_l5_bias_q$plot
dev.off()
```

### Table S3
```{r}
test_weighting_vs_bias_batch(xubsh_diff_qnorm)
```

## Load up NSC / Neuron data
### Figure S8 is taken from the occupancy diagnostic plots
```{r}
nvsn = load_data_genes(binding_profiles_path = c("data/RNAPII_TaDa/*ph.gz"),
                       norm_method = "quantile",
                       occupancy_plot_diagnostics = T)
```

```{r}
nvsn.diff = differential_binding(nvsn, cond=c("NSCs"="Wor","Neurons"="elav"))
```

### This figure didn't make it into the manuscript, but kinda neat regardless if you care for Drosophila neural dev biol :)
```{r}
go.nscs = analyse_go_enrichment(nvsn.diff,direction="NSCs",maxGSSize = 500,show_category = 18, label_format_width = 60, fit_labels = F)
go.neurons = analyse_go_enrichment(nvsn.diff,direction="Neurons",maxGSSize = 500,show_category = 18, label_format_width = 60, fit_labels = F)

getgo = function (go_analyis,go_id) {stringr::str_split(go_analyis$results_table[go_id,"geneID"],"/") |> unlist()}
syn_signalling = getgo(go.neurons,"GO:0099536")
neuroblast_diff = getgo(go.nscs,"GO:0014016")

plot_volcano(nvsn.diff, 
             label_config=NULL, 
             highlight = list(
                "Neuroblast differentiation"=neuroblast_diff, 
                "Synaptic signalling"=syn_signalling
                ), 
             highlight_config = list(max_overlaps=15, clean_names=T),
             fdr_filter_threshold = 0.01)
```
